Best Practices in Mixture Modeling using Free Open Source Software

### 2022-09-02 manuscript.Rmd
Updates 02 September
Included relevant info from comments up to and including "Use cases for latent class analysis"

Wrote examples for "Use cases for latent class analysis".

Latent class analysis (LCA) is an umbrella term that refers to a number of techniques for estimating unobserved group membership
based on a parametric model of one or more observed indicators of group membership.
This method has become quite popular across scientific fields, and under a number of different names; most notably (finite Gaussian) mixture modeling and latent profile analysis.
<!-- Vermunt and Magidson (2004) defined it more generally as virtually any statistical model where “some of the parameters … differ across unobserved subgroups” (Vermunt, J.K. and Magidson, J., 2004. Latent class analysis. In: M.S. Lewis-Beck, A. Bryman, and T.F. Liao (eds.), The Sage Encyclopedia of Social Sciences Research Methods, 549-553. Thousand Oaks, CA: Sage Publications). -->
Despite the popularity of the method, there is a lack of standards for estimating and reporting latent class analyses.
<!-- Prior work resulted in the "GRoLTS-Checklist"; reporting guidelines for a specific subcategory of latent class analysis known as latent growth models. vermuntj.k.LatentClassAnalysis2004-->
<!-- General reporting guidelines for latent class analysis are still lacking, however. -->
This complicates manuscript review and assessment of the quality of published studies,
and introduces a risk of misapplications of the technique.
The present paper seeks to address this gap in the literature by suggesting updated guidelines for estimation and reporting on latent class analysis, based on current best practice.
Importantly, in order to lower the barrier of entry and ensure reproducibility of all examples, this paper exclusively relies on free, open source software for latent class analysis in R.
Our goal is to make best-practices latent class analysis widely accessible.

@@

Latent class analysis can be understood as a method for estimating unobserved groups
based on a parametric model of observed indicators of group membership.
The concept of latent class analysis can be understood in different ways.
Generally speaking, it can be said that a mixture model assumes that the study population
is composed of $K$ subpopulations, or classes.
It further assumes that the observed data are a mixture of data generated by class-specific models.
The simplest univariate "model" is a normal distribution, which can be described with two parameters:
the mean and the variance.
Commonly, the same model is estimated across all classes, but with different parameters for each class (i.e., class-specific means and variances).
Mixture modeling then estimates both the parameters for each class, and the probability for each that an individual belongs to each class.

@@

One way to is to conceptualize latent class analysis is by analogy to a measurement model in structural equation modeling.
A mixture model is like confirmatory factor analysis, except that the continuous latent variable is substituted with a categorical latent variable.
One difference between the two techniques is that factor analysis can be considered as a way to group observed *variables* into latent constructs, with factor loadings indicating which items belong are most indicative of a construct.
By contrast, mixture modeling groups *individuals* into classes [see @nylund-gibson_ten_2018].
In line with this distinction, latent class analysis is sometimes referred to as a "person-centered" technique, and factor analysis as a "variable-centered" technique.

When the focus is on the model parameters in each group,
then latent class analysis can be thought of as similar to a multi-group structural equation model.
The main distinction is that group membership is not known a-priori,
but is instead estimated - with measurement error - based on the data.
Whereas in a multigroup model, the data are split by group and treated as independent samples,
in a mixture model, all cases contribute to the estimation of all parameters in all groups.
The relative contribution of each case to the parameters of each group is determined by that case's posterior probability of belonging to that group.

@@

One common type of latent class analyses is the *finite Gaussian mixture model*;
a univariate analysis where the observed distribution of a single variable is assumed to result from a mixture of a known number of Gaussian (normal) distributions.
The parameters of a finite Gaussian mixture model are the means and variances of these underlying normal distributions.
The analysis of shoe sizes presented earlier is a canonical example of this type of analysis.
In the multivatiate case, with more than one indicator variable,
the parameters of a mixture model are the means, variances, and covariances between the indicators (which can be standardized to obtain correlations).
These parameters can be estimated freely, or constrained, across classes.

@@

The term latent class analysis originally referred to cases where the observed indicators were categorical.
Nowadays, it is more commonly used as an umbrella term.
To prevent ambiguity, the special case where indicators are of binary or ordinal measurement level might be described as latent class analysis with ordinal indicators.
Latent class models with ordinal indicators are parameterized differently from mixture models.
One common parameterization assumes that each categorical variable reflects an underlying standard normal distribution.
The parameters are "thresholds" that correspond to quantiles of a standard normal distribution (with $N(\mu = 0, \sigma = 1)$).
These thresholds are estimated based on the proportion of individuals in each of the response categories of the indicator variable.
For example, a binary indicator has a single threshold that distinguishes the two response categories.
If responses are distributed 50/50, then the corresponding threshold would be  $t_1 = 0.00$.
If the responses are distributed 60/40, then the resulting threshold would be $t_1 = 0.25$.
This paper will primarily focus on mixture models and special cases thereof,
although most of the suggested guidelines are applicable to latent class analyses.

<!-- If a multivariate mixture estimation is constrained so that measures must be uncorrelated within each distribution it is termed latent profile analysis. Modified to handle discrete data, this constrained analysis is known as LCA. Discrete latent trait models further constrain the classes to form from segments of a single dimension: essentially allocating members to classes on that dimension: an example would be assigning cases to social classes on a dimension of ability or merit. -->

@@

LCA can be used as a descriptive analysis where a researcher wishes to describe their sample 
and identify a few prototypes based on many variables.
<!-- add an example and elaborate -->

With LCA, our goal could be to inductively identify the number of classes. 
For instance, if we believe that a variable represents a group, 
but don't know how many groups there are, LCA may be an appropriate technique to answer this question.
<!-- add an example -->

Another application of LCA is to classify individuals.
<!-- add an example and elaborate -->

LCA is also appropriate when we wish to identify indicators that capture classes well.
<!-- add an example and elaborate -->

An extension of LCA is that containing covariates which can be used to predict class membership.
<!-- add an example and elaborate -->

When our interest is the prediction of one or more outcomes,
LCA can be used to construct classes as categorical predictors.
<!-- add an example and elaborate -->

In addition to these applications, LCA can be used for dimensionality reduction
as the resulting groups summarize response patterns on a large number of indicators.
<!-- add an example -->

Finally, LCA can be used to deal with data which violate the normality assumption.
<!-- add an example and elaborate, check with Caspar if there are other assumptions -->

<!-- CJ: Include these use cases too: -->
<!-- Descriptive analysis; identify few prototypes based on many variables, e.g. DJA science funding -->
<!-- Inductively identify number of latent classes: I know it's a group, but not how many groups -->
<!-- Classify individuals, e.g. welzijnsmeter -->
<!-- Identify items (observed characteristics) that -->
<!-- indicate these classes well -->
<!-- Predict class membership from covariates -->
<!-- Predict outcomes from class membership -->
<!-- Dimension reduction -->
<!-- Deal with violation of assumptions, e.g. when data are clearly not normally distributed -->

>>>>

Latent class analysis (LCA) is an umbrella term that refers to a number of techniques for estimating unobserved group membership
based on a parametric model of one or more observed indicators of group membership.
This method has become quite popular across scientific fields, and under a number of different names;
most notably (finite Gaussian) mixture modeling and latent profile analysis.
@vermunt_jk_latent_2004 defined it more generally as virtually any statistical model where “some of the parameters […] differ across unobserved subgroups”.

Despite the popularity of the method, there is a lack of standards for estimating and reporting latent class analyses.
While @van_de_schoot_grolts-checklist_2017 developed reporting guidelines for a specific subcategory of LCA known as latent growth models,
general reporting guidelines for latent class analysis are still lacking.
This complicates manuscript review and assessment of the quality of published studies,
and introduces a risk of misapplications of the technique.
The present paper seeks to address this gap in the literature by suggesting updated guidelines for estimation and reporting on latent class analysis, based on current best practice.
Importantly, in order to lower the barrier of entry and ensure reproducibility of all examples, 
this paper exclusively relies on free, open source software for latent class analysis in R.
Our goal is to make best-practices in latent class analysis widely accessible.

@@

Latent class analysis can be understood as a method for estimating unobserved groups
based on a parametric model of observed indicators of group membership.
The concept of latent class analysis can be understood in different ways.
Generally speaking, a mixture model assumes that the study population
is composed of $K$ subpopulations or classes.
It further assumes that the observed data are a mixture of data generated by class-specific models.
The simplest univariate "model" is a normal distribution, which can be described with two parameters:
the mean and the variance.
Commonly, the same model is estimated across all classes, but with different parameters for each class (i.e., class-specific means and variances).
Mixture modeling then estimates both the parameters for each class, 
and the probability that an individual belongs to each class.

@@

A mixture model is like confirmatory factor analysis, except that the continuous latent variable is substituted with a categorical latent variable.
One difference between the two techniques is that factor analysis can be considered as a way to group observed *variables* into latent constructs, with factor loadings indicating which items belong are most indicative of a construct.
By contrast, mixture modeling groups *individuals* into classes [see @nylund-gibson_ten_2018].
In line with this distinction, latent class analysis is sometimes referred to as a "person-centered" technique, and factor analysis as a "variable-centered" technique.

When the focus is on the model parameters in each group,
then LCA can be thought of as similar to a multi-group structural equation model.
The main distinction is that group membership is not known a-priori,
but is instead estimated - with measurement error - based on the data.
Whereas in a multi-group model, the data are split by group and treated as independent samples,
in a mixture model, all cases contribute to the estimation of all parameters in all groups.
The relative contribution of each case to the parameters of each group is determined by that case's posterior probability of belonging to that group.

@@

One common type of LCA is the *finite Gaussian mixture model*;
a univariate analysis where the observed distribution of a single variable is assumed to result from a mixture of a known number of Gaussian (normal) distributions.
The parameters of a finite Gaussian mixture model are the means and variances of these underlying normal distributions.
The analysis of shoe sizes presented earlier is a canonical example of this type of analysis.
In the multivatiate case, with more than one indicator variable,
the parameters of a mixture model are the means, variances, and covariances between the indicators (which can be standardized to obtain correlations).
These parameters can be estimated freely, or set to be constrained across classes.

@@

The term latent class analysis originally referred to cases where the observed indicators were categorical.
Nowadays, it is more commonly used as an umbrella term.
To prevent ambiguity, the special case where indicators are of binary or ordinal measurement level might be described as *latent class analysis with ordinal indicators*.
Latent class models with ordinal indicators are parameterized differently from mixture models.
One common parameterization assumes that each categorical variable reflects an underlying standard normal distribution.
The parameters are "thresholds" that correspond to quantiles of a standard normal distribution (with $N(\mu = 0, \sigma = 1)$).
These thresholds are estimated based on the proportion of individuals in each of the response categories of the indicator variable.
For example, a binary indicator has a single threshold that distinguishes the two response categories.
If responses are distributed 50/50, then the corresponding threshold would be  $t_1 = 0.00$.
If the responses are distributed 60/40, then the resulting threshold would be $t_1 = 0.25$.

This paper will primarily focus on mixture models and special cases thereof,
although most of the suggested guidelines are applicable to latent class analyses.

<!-- DA: Do we still need the following comment? In my opinion the included text is sufficient. -->
<!-- If a multivariate mixture estimation is constrained so that measures must be uncorrelated within each distribution it is termed latent profile analysis. Modified to handle discrete data, this constrained analysis is known as LCA. Discrete latent trait models further constrain the classes to form from segments of a single dimension: essentially allocating members to classes on that dimension: an example would be assigning cases to social classes on a dimension of ability or merit. -->

@@

LCA can be used as a descriptive analysis where a researcher wishes to describe their sample and identify a few prototypes based on many variables.
As an example, if a survey among all Dutch academics was carried out
with the goal of bringing about a funding reform, 
LCA could be used to discover the types of publications that get funding.

With LCA, our goal could be to inductively identify the number of classes. 
For instance, if we believe that a variable represents a group, 
but don't know how many groups there are, LCA may be an appropriate technique to answer this question.
For example, @hopfer_social_2014 studied the substance use, sexual behavior, and mental health status of urban population in Winnipeg, Canada. 
The underlying assumption was that there were different risk profiles,
but their number was not known.
From a collection of indicators, the LCA provided evidence that there may be four distinct risk profiles in the Winnipeg area.

Another application of LCA is to classify individuals.
In a peer harassment study, @giang_using_2008 used latent class analysis to classify over 2,000 sixth grade students into aggressor and victim latent classes.
The five-class solution comprised of classes of victims, aggressors, and socially adjusted students. For instance, it revealed that there were two types of victims: highly-victimized aggressive-victims and highly-aggressive aggressive-victims.

LCA is also appropriate when we wish to identify indicators that capture classes well.
High quality indicators are strongly related to the latent variable
and lead to good class separation.
This relationship of high quality indicators to the latent variable
is reflected in very high or very low conditional response probabilities.
As a rule of thumb, high indicator quality can be indicated 
by conditional response probabilities above .9 and below .1. 
Moderate indicator quality can be indicated by probabilities above .8 and below .2,
and low indicator quality would be above .7 and below .3.
For a simulation study exploring the effects of indicator quality on LCA, see @geiser_is_2014.
Therefore, one could use conditional response probabilities for each item to assess its quality
with regards to how well it helps separate the latent classes.
From this, a theory about the selection of indicators could be informed.

<!-- "However, Galindo-Garre and Vermunt (2006) found that indicators with high population values (i.e., conditional response probabilities close to one) were more likely to produce boundary estimates in LCA applications, even though this effect decreased with larger N. Still, there is evidence that using high quality indicators is generally beneficial, at least in the context of structural equation models with continuous latent variables (Marsh et al., 1998). Having more high quality indicators can stabilize the estimation by increasing the information available to estimate latent variable parameters. In this vein, Collins and Wugalter (1992) speculated that having sufficiently high quality indicators may compensate for having few indicators and may aid parameter recovery in LTA models." direct quote from @geiser_is_2014 -->

An extension of LCA is that containing covariates which can be used to predict class membership.
In this approach, we not only model the latent class variable based on indicators,
but we also relate the class membership to other explanatory variables (@vermunt_latent_2017).
An example of using covariates comes from @nozadi_moderating_2016 
who applied LCA to identify the probability of children's membership to an anxiety class.
The authors tested several covariates including children's age, sex, and accuracy scores.
Age and sex were not found to be related to the children's latent class membership,
hence these covariates were excluded from the analysis.
Accuracy scores were related to probabilities of being in anxiety and attention-
anxiety classes and therefore this covariate was kept as a valuable predictor.

When our interest is the prediction of one or more outcomes,
LCA can be used to construct latent classes as categorical predictors.
@lanza_latent_2013 demonstrated how LCA can be used to classify adolescents into depression classes,
and subsequently these classes can be used to predict smoking, grades, and delinquency.
The study showed that the outcomes predicted by class membership can be binary (regular smoking), 
continuous (grades) or count (delinquency).

In addition to these applications, LCA can be used for dimensionality reduction
as the resulting groups summarize response patterns on a large number of indicators.
For example, @macgregor_symptom_2021 investigated symptom profiles among injured U.S. military personnel. 
They used fifteen dichotomous items from the Post-Deployment Health Assessment survey as LCA indicators.
Combinatorics informs us that fifteen dichotomous items have $2^{15}$ or $32,768$ unique symptom combinations.
Perhaps for this reason, @macgregor_symptom_2021 incorporated LCA as a method of dimensionality reduction.
A five class solution was found to have the best fit according to both statistical criteria and clinical interpretability.

<!-- from @macgregor_symptom_2021 "Incorporating LCA with the symptom checklist was a novel method of dimensionality -->
<!-- reduction, and the identified symptom profiles provided more granularity than examining individual symptoms -->
<!-- alone. There are also limitations that warrant mention. At times, LCA can be prone to classification errors, but in the present analysis, the model entropy value above 0.80 indicated good separation of the classes, and the MeanPP values showed a low rate of misclassification." -->
<!-- [...] -->
<!-- [listing the five classes] "(1) 50.4% were low -->
<!-- morbidity (i.e., no or very few symptoms reported), (2) -->
<!-- 14.0% musculoskeletal (i.e., muscle aches; swollen, stiff or -->
<!-- painful joints; back pain), (3) 8.9% auditory (i.e., trouble -->
<!-- hearing; ringing in the ears), (4) 11.1% psycho-cognitive -->
<!-- (i.e., trouble concentrating, easily distracted; forgetful or -->
<!-- trouble remembering things; increased irritability), and (5) -->
<!-- 15.6% multimorbidity (i.e., multiple symptoms reported)." -->

Finally, LCA can be used to deal with data which violate certain assumptions.
For instance, nested data usually imply a violation of the independence assumption.
In this case, an extension of LCA called multilevel LCA can be used (e.g. @henry_multilevel_2010).
As discussed in the shoe size example, LCA can deal with violations of normality.
In fact, LCA assumes the population distribution in a non-normal mixture of $K$ normal distributions,
and LCA can discover the value of $K$, i.e. generate a $K$-class solution.

<!-- check with Caspar if there are other assumptions and how to expand this paragraph -->

<!-- OLD: -->
<!-- CJ: Include these use cases too: -->
<!-- Descriptive analysis; identify few prototypes based on many variables, e.g. DJA science funding -->
<!-- Inductively identify number of latent classes: I know it's a group, but not how many groups -->
<!-- Classify individuals, e.g. welzijnsmeter -->
<!-- Identify items (observed characteristics) that -->
<!-- indicate these classes well -->
<!-- Predict class membership from covariates -->
<!-- Predict outcomes from class membership -->
<!-- Dimension reduction -->
<!-- Deal with violation of assumptions, e.g. when data are clearly not normally distributed -->
